### 2022-11-15 chapter1.tex
Re-wrote various captions and figure references to make them sound more like a journal

\begin{figure}[h!]
\includegraphics[width=\textwidth]{graphical_overview.png}
\centering
\caption{\textbf{Graphical overview of homology inference pipeline.}
In panels A-C, the large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation. Sequences that share homology are colored with shades of the same hue. In panels D-F, the small, filled circles and lines represent the same sequences and best hits from the previous steps.}
\label{fig:graphical_overview}
\end{figure}

\section*{Results}
\subsection*{Pipeline overview}
The orthology inference pipeline is summarized in Figure \ref{fig:graphical_overview}. Overall, the approach is similar to other graph-based methods of orthology inference. First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are then conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B). The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C). Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D). Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E). Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F). In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.

@@

\subsection*{Clustering in orthologous groups}
The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes. Two connected components, sets of nodes joined by a sequence of edges, are shown in Figure \ref{fig:components}. The sequences in panel A all contain C2H2 zinc fingers whereas the sequences in panel B are members of the Par-1 family of serine/threonine protein kinases. In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors. To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each. The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively. We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C). Two distinct trendlines are apparent. The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene. The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes. Thus, there are generally two classes of components. The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{components/out/0017-005F.png}
\centering
\caption{\textbf{Selected connected components of hit graph.}
Panels A and B are two distinct connected components of the graph formed by the best hits between sequences. Edges are colored by the value of their bit score.}
\label{fig:components}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{percolation/percolation.png}
\includegraphics[width=\textwidth]{percolation/out/005F.png}
\centering
\caption{\textbf{Clique percolation algorithm.}
A) A \textit{k}-clique is a set of \textit{k} nodes which are fully connected, i.e., there is an edge between pair of nodes. Cliques are shown for \textit{k} equal to three, four, five, and six. \textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively. B) The clique percolation algorithm detects clusters by “rolling” a clique of some size \textit{k} over the graph. More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified. The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique. A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster. In this example, \textit{k} was set to four. C) A single component clustered by clique percolation with varying values of \textit{k} is shown. Larger values of \textit{k} define tighter clusters. Nodes are colored according to their cluster. If a node belongs to multiple clusters, it uses a blend of those colors.}
\label{fig:percolation}
\end{figure}

@@

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions. Thus, in our pipeline the sequences were aligned with a conservative value of 0.4. Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors. To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}. The original sequences were then aligned to these profiles. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments. Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini. This effect is visible in the three examples shown in Figure \ref{fig:realignment}, where each has long poorly supported segments that are completely removed.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{realignment/out/merged.png}
\centering
\caption{\textbf{Alignment with long poorly supported segments.}
The alignments of representative sequences in orthologous groups 23D9 (A), 01E0 (B), and 07C3 (C) are shown before and after refinement.}
\label{fig:realignment}
\end{figure}

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states” \cite{Felsenstein1996}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively. The probability distributions for each state shared a common structure, which is illustrated in Figure \ref{fig:hmm}A. We conceptualized the observed alignments as the superposition of two distinct processes. The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by "noise" from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we first coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences. The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the "gap stickiness" as a beta-binomial random variable by counting the number of symbols that remain constant between columns. The parameters associated with each component are shown to the right of Figure \ref{fig:hmm}A. Each state has a different set of parameters that yield its characteristic gap pattern and gap stickiness.

After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment. Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. An example decoded alignment is shown in Figure \ref{fig:hmm})B where the decoded states closely follow the expected patterns. Overall, fewer than 10\% of alignments were trimmed of any regions. However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:hmm_trim}).

\begin{figure}[h!]
\includegraphics[width=\textwidth]{hmm/hmm_architecture.png}
\includegraphics[width=\textwidth]{hmm/out/2252.png}
\centering
\caption{\textbf{HMM emission architecture and a decoded alignment.}
A) White and colored boxes indicate gap and non-gap symbols in the biological sequences, respectively. White and grey boxes indicate gap and non-gap symbols in the encoded sequences, respectively. c\textsubscript{0} and c\textsubscript{1} indicate the first and second columns in the alignment, respectively. B) The alignment of the representative sequences in orthologous group 2252 is decoded using the trained HMM.}
\label{fig:hmm}
\end{figure}

@@

\subsection*{Inference of species trees}
Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences. In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition. Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences. However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}. Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups. We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A). To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were "reverse translated" from the protein alignments and their corresponding coding sequences. The resulting consensus tree is shown in Figure \ref{fig:trees}B. Both trees have an identical topology, which is consistent with other published phylogenies \cite{D12GC2007, DaLage2007}.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{trees/out/trees1.png}
\centering
\caption{\textbf{Phylogenetic tree of species.}
A) The LG model was fit to meta-alignments directly sampled from the original protein alignments B). The GTR model was fit to meta-alignments sampled from "reverse translated" nucleotide alignments. Values at nodes are bootstrap percentages.}
\label{fig:trees}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{clusters/out/clusters.png}
\centering
\caption{\textbf{Statistics of orthologous groups.}
A) Each species is equally represented in orthologous groups (OGs). B) A plurality of orthologous groups contain all species. C) Nearly all proteins are associated with a single orthologous group. D) The number of orthologous groups associated with a species is strongly correlated with the number of unique annotated proteins, which suggests the annotation pipeline generally identifies conserved genes.}
\label{sfig:clusters}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{paralogs/out/paralogs.png}
\centering
\caption{\textbf{Addition of paralogs to orthologous groups.}
A) Most orthologous groups (OGs) have no in-paralogs. B, D) Of the groups with paralogs, most have fewer than five. C) The in-paralogs are generally only a small fraction of the sequences in an orthologous group.}
\label{sfig:paralogs}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{cnn_training/out/cnn_training.png}
\includegraphics[width=\textwidth]{cnn_training/out/23D9_XP_026832050.1.png}
\centering
\caption{\textbf{CNN data and training details.}
A) Most labels were “negative,” which corresponded to positions which were maintained in the alignment. B, C) The training loss and metrics stabilized by the final training epochs. D, E) The parameters for a model of this size are easily visualized. The embedding learns the gap and unknown amino acid symbols are distinct from the others. The convolutional filters learn kernels which correspond to the length of a typical long poorly supported segment. F) The alignment of representative sequences in orthologous group 23D9 where the indicated sequence is labeled with the trained CNN. The blue lines are the CNN’s predictions, and the yellow lines are the labels. The red lines are the weights given to each position. The regions where the weights are zero are masked from the training data, which allowed the model to focus on only the most relevant examples.}
\label{sfig:cnn_training}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{hmm_training/out/hmm_training.png}
\centering
\caption{\textbf{Insertion phylo-HMM data and training details.}
A) Most columns in the training data were labeled as state 1A or 1B. B-F) The model loss and parameters largely stabilized by the final training iteration. Panels C-F correspond to the parameters in the phylogenetic process, the jump process, the pattern stickiness model, and the transition matrix, respectively. The transition matrix plots are the transition rates to the state indicated on the vertical axis and given in log scale. Self transitions are excluded.}
\label{sfig:hmm_training}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{hmm_trim/out/hmm_trim.png}
\centering
\caption{\textbf{Insertion phylo-HMM trimming details.}
A) Most alignments were not trimmed of any columns. B) Most alignments have fewer than 5 trimmed regions which cumulatively account for less than 10\% of the columns in the original alignment. Alignments with more trimmed columns generally have one one large trimmed region rather than many small ones. C) Most trimmed regions were inferred as state 3.}
\label{sfig:hmm_trim}
\end{figure}

>>>>

\begin{figure}[h!]
\includegraphics[width=\textwidth]{graphical_overview.png}
\centering
\caption{\textbf{Overview of homology inference pipeline.}
\textbf{(A-C)} The large, open circles represent the annotated genomes, and the small, filled circles represent the protein sequences associated with each annotation. Sequences that share homology are colored with shades of the same hue. \textbf{(D-F)} The small, filled circles and lines represent the same sequences and best hits from the previous steps.}
\label{fig:graphical_overview}
\end{figure}

\section*{Results}
\subsection*{Pipeline overview}
Our pipeline follows a similar overall approach to other graph-based methods of orthology inference. First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are then conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B). The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C). Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D). Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E). Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F). In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.

@@

\subsection*{Clustering in orthologous groups}
The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes. Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A,B). The sequences in the first (Fig.~\ref{fig:components}A) all contain C2H2 zinc fingers, whereas the sequences in the second (Fig.~\ref{fig:components}B) are members of the Par-1 family of serine/threonine protein kinases. In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors. To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each. The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively. We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C). Two distinct trendlines are apparent. The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene. The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes. Thus, there are generally two classes of components. The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{components/out/0017-005F.png}
\centering
\caption{\textbf{Selected connected components of hit graph and summary statistics.}
\textbf{(A-B)} Two distinct connected components of the hit graph. Edges are colored by the value of their bit score. \textbf{(C)} Hexbin plot of the number of sequences and the number of unique genes in each component. \textbf{(D)} Histogram of number edges associated with each sequence, \textit{i.e.} the degree of each node.}
\label{fig:components}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{percolation/percolation.png}
\includegraphics[width=\textwidth]{percolation/out/005F.png}
\centering
\caption{\textbf{Clique percolation algorithm.}
\textbf{(A)} A \textit{k}-clique is a set of \textit{k} nodes which are fully connected, i.e., there is an edge between pair of nodes. Cliques are shown for \textit{k} equal to three, four, five, and six. \textit{k} equal to one and two correspond to a single node and two nodes joined by an edge, respectively. \textbf{(B)} The clique percolation algorithm detects clusters by “rolling” a clique of some size \textit{k} over the graph. More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified. The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique. A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster. In this example, \textit{k} was set to four. \textbf{(C)} A single component clustered by clique percolation with varying values of \textit{k} is shown. Larger values of \textit{k} define tighter clusters. Nodes are colored according to their cluster. If a node belongs to multiple clusters, it uses a blend of those colors.}
\label{fig:percolation}
\end{figure}

@@

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions. Thus, in our pipeline the sequences were aligned with a conservative value of 0.4. Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors. To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}. The original sequences were then aligned to these profiles. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments. Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini. This effect is visible in the illustrated examples, where each has long poorly supported segments that are completely removed (Fig.~\ref{fig:realignment}).

\begin{figure}[h!]
\includegraphics[width=\textwidth]{realignment/out/merged.png}
\centering
\caption{\textbf{Alignment with long poorly supported segments.}
The alignments of representative sequences in orthologous groups 23D9 \textbf{(A)}, 01E0 \textbf{(B)}, and 07C3 \textbf{(C)} before and after refinement.}
\label{fig:realignment}
\end{figure}

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states” \cite{Felsenstein1996}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by "noise" from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the "gap stickiness" as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment. Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~ref{fig:hmm})B). Overall, fewer than 10\% of alignments were trimmed of any regions. However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:hmm_trim}).

\begin{figure}[h!]
\includegraphics[width=\textwidth]{hmm/hmm_architecture.png}
\includegraphics[width=\textwidth]{hmm/out/2252.png}
\centering
\caption{\textbf{HMM emission architecture and a decoded alignment.}
\textbf{(A)} Schematic of theoretical alignment generating process and corresponding probabilistic components in HMM. White and colored boxes indicate gap and non-gap symbols in the biological sequences, respectively. White and grey boxes indicate gap and non-gap symbols in the encoded sequences, respectively. c\textsubscript{0} and c\textsubscript{1} indicate the first and second columns in the alignment, respectively. The parameters associated with each component are shown to the right. \textbf{(B)} The alignment of the representative sequences in orthologous group 2252 decoded using the trained HMM.}
\label{fig:hmm}
\end{figure}

@@

\subsection*{Inference of species trees}
Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences. In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition. Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences. However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}. Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups. We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A). To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were "reverse translated" from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B). Both trees have an identical topology, which is consistent with other published phylogenies \cite{D12GC2007, DaLage2007}.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{trees/out/trees1.png}
\centering
\caption{\textbf{Phylogenetic tree of species.}
\textbf{(A)} Consensus tree from LG model fit to meta-alignments directly sampled from the original protein alignments. \textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from "reverse translated" nucleotide alignments. Values at nodes are bootstrap percentages.}
\label{fig:trees}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{clusters/out/clusters.png}
\centering
\caption{\textbf{Statistics of orthologous groups.}
\textbf{(A)} Each species is equally represented in orthologous groups (OGs). \textbf{(B)} A plurality of orthologous groups contain all species. \textbf{(C)} Nearly all proteins are associated with a single orthologous group. \textbf{(D)} The number of orthologous groups associated with a species is strongly correlated with the number of unique annotated proteins, which suggests the annotation pipeline generally identifies conserved genes.}
\label{sfig:clusters}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{paralogs/out/paralogs.png}
\centering
\caption{\textbf{Addition of paralogs to orthologous groups.}
\textbf{(A)} Most orthologous groups (OGs) have no in-paralogs. \textbf{(B, D)} Of the groups with paralogs, most have fewer than five. \textbf{(C)} The in-paralogs are generally only a small fraction of the sequences in an orthologous group.}
\label{sfig:paralogs}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{cnn_training/out/cnn_training.png}
\includegraphics[width=\textwidth]{cnn_training/out/23D9_XP_026832050.1.png}
\centering
\caption{\textbf{CNN data and training details.}
\textbf{(A)} Most labels were “negative,” which corresponded to positions which were maintained in the alignment. \textbf{(B, C)} The training loss and metrics stabilized by the final training epochs. \textbf{(D, E)} The parameters for a model of this size are easily visualized. The embedding learns the gap and unknown amino acid symbols are distinct from the others. The convolutional filters learn kernels which correspond to the length of a typical long poorly supported segment. \textbf{(F)} The alignment of representative sequences in orthologous group 23D9 where the indicated sequence is labeled with the trained CNN. The blue lines are the CNN’s predictions, and the yellow lines are the labels. The red lines are the weights given to each position. The regions where the weights are zero are masked from the training data, which allowed the model to focus on only the most relevant examples.}
\label{sfig:cnn_training}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{hmm_training/out/hmm_training.png}
\centering
\caption{\textbf{Insertion phylo-HMM data and training details.}
\textbf{(A)} Most columns in the training data were labeled as state 1A or 1B. \textbf{(B)} The model loss stabilized by the final training iteration. \textbf{(C-F)} The values of parameters in the phylogenetic process, the jump process, the pattern stickiness model, and the transition matrix, respectively, at each training iteration. The transition matrix plots are the transition rates to the state indicated on the vertical axis and given in log scale. Self transitions are excluded.}
\label{sfig:hmm_training}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{hmm_trim/out/hmm_trim.png}
\centering
\caption{\textbf{Insertion phylo-HMM trimming details.}
\textbf{(A)} Most alignments were not trimmed of any columns. \textbf{(B)} Most alignments have fewer than 5 trimmed regions which cumulatively account for less than 10\% of the columns in the original alignment. Alignments with more trimmed columns generally have one one large trimmed region rather than many small ones. \textbf{(C)} Most trimmed regions were inferred as state 3.}
\label{sfig:hmm_trim}
\end{figure}
